Type Classes for Mathematics in Type Theory

### 2010-12-09 papers/mscs/mscs.tex
Still more referee-guided improvements in MSCS paper.

\paragraph{Outline.}
In section~\ref{preliminaries} we briefly describe the Coq system and its implementation of type classes. Then, in section~\ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, arguably the biggest design dimension when building interfaces for abstract structures. In section~\ref{predicateclasses} we show how type classes can make the use of `unbundled' purely predicate based interfaces for abstract structures practical. Next, in section~\ref{classes}, we discuss our algebraic hierarchy implemented using such predicate classes. In sections~\ref{cat} and~\ref{univ} we give a taste of what category theory and universal algebra look like in our development, and in section~\ref{numbers} we use these facilities to build abstract interfaces for numbers. In order to illustrate a very different use of type classes, we discuss the implementation of a quoting function for algebraic terms in terms of type classes, in section~\ref{quoting}. In section~\ref{sequences} we hint at an interface for sequences, but describe how a limitation in the Coq system makes its use problematic. We end with conclusions in section~\ref{conclusions}.

@@

% \paragraph{Equality, setoids, and rewriting}

The `native' notion of equality in Coq is that of terms being convertible, naturally reified as a proposition by the inductive type family \lstinline|eq| with single constructor \lstinline|eq_refl|:
\begin{lstlisting}
  eq_refl : ∀ (T: Type) (x: T), x ≡ x,
\end{lstlisting}
where `\lstinline|a ≡ b|' is notation for \lstinline|eq T a b|. Here we diverge from Coq tradition and reserve \lstinline|=| for setoid equality, as this is the equality we will be working with most of the time. Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions. We mention this explicitly only because rewriting \emph{does} give rise to conditions when we depart from Leibniz equality and introduce equivalence relations representing abstractions over common representations of the same conceptual object. For instance, consider formal fractions of integers as a representation of rationals. Rewriting a subterm using such an equality is permitted only if the subterm is an argument of a function that has been proven to \emph{respect} the equality. Such a function is called \emph{proper}, and propriety must be proved for each function in whose arguments we wish to enable rewriting.

@@

The rationals are characterized as a decidable field with an injective ring morphism from a canonical implementation of the integers and a surjection of fractions of such integers:
\begin{lstlisting}
  Context `{Field A} `{∀ x y: A, Decision (x = y)} {inj_inv}.
  Class Rationals: Prop :=
    { rationals_field:> Field A
    ; rationals_frac: Surjective
        (λ p => integers_to_ring (Z nat) A (fst p) *
          / integers_to_ring (Z nat) A (snd p)) (inv:=inj_inv)
    ; rationals_embed_ints: Injective (integers_to_ring (Z nat) A) }.
\end{lstlisting}
Here, \lstinline|Z| is an \lstinline|Integers| implementation paramerized by a \lstinline|Naturals| implementation, for which we just take \lstinline|nat|. The choice of \lstinline|Z nat| here is immaterial; we could have picked another, or even a generic, implementation of \lstinline|Integers|, but doing so would provide no benefit.
  % todo: hm, this Z is poorly named. it's really natpair_integers.Z
  % todo: say a few words about how we handle division by zero?

@@

Universe polymorphism could in principle most likely solve this problem, but its current implementation in Coq only supports universe polymorphic inductive definitions, while \lstinline|Singleton| is a regular definition. Universe polymorphic \emph{definitions} have historically not been supported in Coq, primarily because of efficiency concerns. However, we have taken up the issue with the Coq development team, and they have agreed to introduce the means to let one turn on universe polymorphism for definitions voluntarily on a per-definition basis. With such functionality we could make \lstinline|Singleton| universe polymorphic, and hopefully resolve these problems.

>>>>

\paragraph{Outline.}
In section~\ref{preliminaries} we briefly describe the Coq system and its implementation of type classes. Then, in section~\ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, arguably the biggest design dimension when building interfaces for abstract structures. In section~\ref{predicateclasses} we show how type classes can make the use of `unbundled' purely predicate based interfaces for abstract structures practical.

In the rest of the paper, we tour the key components in our development, leading up to the numerical interfaces. This will not only show the pleasant style of formalization that rigorous use of type classes enables, but will also show that eager adoption and incorporation of more abstract mathematical perspectives (which are traditionally often ignored when doing dependently typed programming on concrete data structures in type theory), is not only feasible but actually practical and beneficial.

 In section~\ref{classes}, we discuss our algebraic hierarchy implemented using type classes. In sections~\ref{cat} and~\ref{univ} we give a taste of what category theory and universal algebra look like in our development, and in section~\ref{numbers} we use these facilities to build abstract interfaces for numbers. In order to illustrate a very different use of type classes, we discuss the implementation of a quoting function for algebraic terms in terms of type classes, in section~\ref{quoting}. In section~\ref{sequences} we hint at an interface for sequences, but describe how a limitation in the current implementation of Coq makes its use problematic. We end with conclusions in section~\ref{conclusions}.

@@

\paragraph{Equality, setoids, and rewriting}

The `native' notion of equality in Coq is that of terms being convertible, naturally reified as a proposition by the inductive type family \lstinline|eq| with single constructor \lstinline|eq_refl|:
\begin{lstlisting}
  eq_refl : ∀ (T: Type) (x: T), x ≡ x,
\end{lstlisting}
where `\lstinline|a ≡ b|' is notation for \lstinline|eq T a b|. Here we diverge from Coq tradition and reserve the \lstinline|=| symbol for \emph{setoid} equality (to be discussed momentarily), as this is the equality we will be working with most of the time. Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions. We mention this explicitly only because such rewriting \emph{does} give rise to conditions when we depart from raw convertibility and introduce equivalence relations
that express how possibly distinct (unconvertible) terms may represent the same conceptual object. A typical example would be different formal integer fractions representing the same rational number. Rewriting a subterm using a proof of such an equality is permitted only if the subterm is argument to a function that has been proven to \emph{respect} the equality. Such a function is called \emph{proper} with respect to the equality in question, and propriety must be proved for each function in whose arguments we wish to enable rewriting.

@@

The rationals are characterized as a decidable field with an injective ring morphism from a canonical implementation of the integers and a surjection of fractions of such integers:
\begin{lstlisting}
  Context `{Field A} `{∀$$x y: A, Decision (x = y)} {inj_inv}.
  Class Rationals: Prop :=
    { rationals_field:> Field A
    ; rationals_frac: Surjective
        (λ p => integers_to_ring (Z nat) A (fst p) *
          / integers_to_ring (Z nat) A (snd p)) (inv:=inj_inv)
    ; rationals_embed_ints: Injective (integers_to_ring (Z nat) A) }.
\end{lstlisting}
Here, \lstinline|Z| is an \lstinline|Integers| implementation paramerized by a \lstinline|Naturals| implementation, for which we just take \lstinline|nat|. The choice of \lstinline|Z nat| here is immaterial; we could have picked another, or even a generic, implementation of \lstinline|Integers|, but doing so would provide no benefit.
  % todo: hm, this Z is poorly named. it's really natpair_integers.Z
  % todo: say a few words about how we handle division by zero?

@@

Universe polymorphism could in principle most likely solve this problem, but its current implementation in Coq only supports universe polymorphic \emph{inductive} definitions, while \lstinline|Singleton| is a regular definition. Universe polymorphic regular definitions have historically not been supported in Coq, primarily because of efficiency concerns. However, we have taken up the issue with the Coq development team, and they have agreed to introduce the means to let one turn on universe polymorphism for definitions voluntarily on a per-definition basis. With such functionality we could make \lstinline|Singleton| universe polymorphic, and hopefully resolve these problems.
